Constructing ordinal outcomes for clinical trials

### 2025-05-29 choosing_ordinal_effect_measures.tex
Simpler names
Begin splitting into two logical manuscripts

\addbibresource{bibliography.bib}
\graphicspath{{fig/}}

@@

>>>>

\addbibresource{references.bib}
\graphicspath{{fig/}}

@@

\section{Desiderata for ordinal effect measures}

What about effect measures? We generally seek effect measures that
\begin{enumerate}
  \item accurately summarize decision-relevant differences in distributions of
    patient benefit and harm
  \item are easy to interpret and to translate into practical recommendations
  \item can be incorporated into statistical models where required
  \item are easy to generalize or ``transport'' across contexts or settings
\end{enumerate}

Key references:
\citet{agrestiMeasuresNominalOrdinalAssociation1981,agrestiOrdinalProbabilityEffect2017a,
agrestiSimpleWaysInterpret2018a}

Goal: we want to compare two or more distributions over ordinal
outcomes with a single number. (Two distributions when comparing control and
experimental groups; more distributions when e.g. adjusting for covariates.)

In general this represents a loss of information. Let's say the
outcome has $C$ categories. Then the distribution in each
arm is specified by $C-1$ numbers (for the first $C-1$ probabilities,
  or cumulative logits, or stopping ratio logits--reflecting the
sum-to-1 constraint on probability distributions) and so the
\emph{difference} between the distributions is also specified by
$C-1$ numbers (for example the differences between the cumulative
logits, or logit stopping probabilities).

An effect measure thus represents a way to collapse these $C-1$
numbers into one.

It usually doesn't make sense to consider effect measures based on
numeric coding (e.g. difference in mean levels). This is adding
information that is not present in the basic ordinal structure.

\section{A taxonomy of ordinal effect measures}

\subsection{Model-based effect measures}

\begin{figure}
  \includegraphics[width=7in]{illustrating_common_or.pdf}
  \caption{The common odds ratio, a model-based effect measure, comes
    from the best-fitting parallel curves to the logit transformed
    empirical cumulative distribution. \textbf{Ideally would illustrate
      using toy bar charts as above, with only 2 groups? Although it's a
  strength that model-based measures can compare multiple groups.}}
  \label{fig:illustrating_or}
\end{figure}

\begin{itemize}
  \item
    common odds ratio (or e.g. probit analogue; there are versions of
    this for any cumulative link)
  \item common stopping ratio or common continuation
    ratio (hazard ratio is an example with implicitly ordinal time)
\end{itemize}

What they have in common: the effect measure is a byproduct of fitting a model
to the data. See Figure \ref{fig:illustrating_or} for an illustration
for the common odds ratio.

A good overview of ordinal models is \citet{burknerOrdinalRegressionModels2019}.

The model assumes that a certain \emph{pattern} is present in the outcome
distributions across groups. If the pattern ``holds'', then the differences in
distributions can be described by a single number (with more than two groups,
$G-1$ numbers) instead of $(C-1)$ numbers.

In Figure \ref{fig:illustrating_or} the pattern assumed to hold is
that the logit-transformed eCDFs in each group are parallel.

\subsection{Model-free or nonparametric effect measures}

These are not based on a model (assumed pattern in the differences
between outcome distributions).

\subsubsection{Effect measures based on ``independent pairs'' or
``win'' constructions}

\begin{figure}
  \includegraphics[width=7.5in]{effect_measures_pp_plot.pdf}
  \caption{Illustrating independent pairwise comparison-based effect
    measures using a PP-plot or ROC curve. Adapted from
  \citet{smithsonReceiverOperatingCharacteristic2023}.}
  \label{fig:illustrating_pairwise_comparisons}
\end{figure}

Probabilistic index/probability of
superiority, win ratio, win odds, etc. These predate win ratio and
DOOR methods
\citep[e.g.][p.~14]{agrestiAnalysisOrdinalCategorical2010}.

Rather they are all be based on the ``\emph{pairs construction}'',
where we imagine independently sampling patients from the control and
treatment groups and considering ``wins'' ``ties'' and ``losses''
(corresponding to concordant and discordant pairs and ties in $y$
respectively). See Figure \ref{fig:illustrating_pairwise_comparisons}
for an illustration.

They can
also be \emph{estimated from data} using pairs of patients, but this
is not essential and they could also be estimated by fitting a
proportional odds model, for example (though they would not be a
simple function of a coefficient).

\subsubsection{Dichotomizing or comparing with a reference group}

\begin{itemize}
  \item Dichotomizing by picking a specific cutpoint,
    e.g. ``hospitalization or
    worse''. This is effectively using a binary outcome, but with the
    ordinal structure used to gain power if fitting a model.
  \item Dichotomizing using quantiles of a reference
    group; for example,
    ``worse than the median outcome under standard care''
  \item Simultaneously comparing all quantiles to those of a
    reference group: ridit analysis
    \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
      smithsonReceiverOperatingCharacteristic2023,
    jansenRiditAnalysisReview1984}
\end{itemize}

\subsection{Comparing model-based and nonparametric effect measures}

Both of these can be considered ``estimands'', insofar as we consider the
estimand associated with the model-based effect measures to be given by the
best-fitting model in the population.

Two classes of assumption violations for model-based effect measures
are \emph{magnitude}-heterogeneity across cutpoints (for example,
  effect of a treatment on hospital length of stay or worse is larger
  than its effect on
death on the relevant scale) and
\emph{sign}-heterogeneity
across cutpoints (failure of stochastic dominance -- for example,
treatment is beneficial for hospital length of stay, harmful for death).

Nonparametric effect measures may appear to be assumption-free. However

\begin{itemize}
  \item Assumptions don't need to hold exactly for the model-based
    effect measures to make sense
  \item In particular, if stochastic dominance holds, we can think of
    both model-based and nonparametric effect measures as
    testing the null hypothesis of no effect ``in a particular
    direction''. A weak interpretation of any particular
    effect measure
    is that it supplies this ``direction''.

    For example the Wilcoxon test is
    equivalent to the score test from fitting a proportional odds model.
    Thus fitting a proportional odds model (or any other model) has a
    valid null-testing
    interpretation even if proportional odds doesn't hold.
  \item Under failure of stochastic dominance (``sign
    heterogeneity''), \emph{neither} parametric nor nonparametric
    effect measures make sense, since it doesn't make sense to
    summarize effects by a single number. Need to dig into the
    cutpoint-specific effects.
  \item Model-free methods don't define a common scale for comparing
    more than two interventions. For example, model-free effect
    measures can be \emph{nontransitive}, so it's not clear how to
    use them in $\geq 2$-arm studies
  \item For the same reason, model-based methods have additional
    advantages, for example
    the possibility of adjustment for covariates.

    When additivity approximately holds, this yields a ``best of both
    worlds'' effect measure that is
    personalized (conditional) yet reportable as a single
    number.
\end{itemize}

Harrell critiques \citep{harrellOverviewCompositeOutcome2024,
harrellRareDegenerativeDiseases2024, harrellViewsCompositeOutcome}:
mainly that
win probability etc. are \emph{relative measures}:

\begin{itemize}
  \item Comparable to Cohen's $d$ or $t$-statistic (difference in
    means divided by standard deviation or standard error) rather than
    difference in means: compare: standardized effect size of 0.2 vs.
    difference in means of 10 mmHg
  \item Affected by study inclusion criteria (why? probably needs to
    be fleshed out.)
\end{itemize}

Simulations show a typically very strong association between
proportional odds ratios and $c$-index
\citep{harrellViolationProportionalOdds2020}.

We should distinguish between effect measures used for \emph{analysis} and for
\emph{reporting}. It may make sense to report the probabilistic index or a
single-cutpoint binary outcome even if the analysis is undertaken using a
proportional odds model.

\newpage

\printbibliography

### 2025-05-29 constructing_ordinal_outcomes.tex
Rename + start effect measures draft
Simpler names
Begin splitting into two logical manuscripts

\addbibresource{bibliography.bib}
\graphicspath{{fig/}}

\title{Guidance for clinicians in constructing ordinal outcomes and choosing
ordinal effect measures for clinical trials}
\author{Leon Di Stefano + Ordinal
Working Group}
\date{\today}

@@

\section{Desiderata for outcomes and effect measures: validity and efficiency}

@@

An example of an outcome which is precise but which may be invalid
might be continuously-measured concentrations of a biomarker which is a poor
surrogate for disease progression.

What about effect measures? We generally seek effect measures that
\begin{enumerate}
  \item accurately summarize decision-relevant differences in distributions of
    patient benefit and harm
  \item are easy to interpret and to translate into practical recommendations
  \item can be incorporated into statistical models where required
  \item are easy to generalize or ``transport'' across contexts or settings
\end{enumerate}

\section{Combining and coarsening ordinal outcomes}

\subsection{3 ways to combine/fine-grain ordinal outcomes: sequential,
hierarchical, \& parallel composition}

@@

\subsubsection{Another way: scores}

@@

\begin{figure}
  \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{Ordinal outcome used in the planned BANDICOOT adaptive
    platform trial \citep{walkerCodesigningNovelOrdinal2025}.
  Illustrates \textbf{parallel composition} of component outcomes}
  \label{fig:parallel_bandicoot}
\end{figure}

\subsection{Reasons to combine ordinal outcomes}

\subsubsection{Incorporating intercurrent events for validity}

@@

\begin{figure}
  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
  \caption{Ordinal outcome used in the REMAP-CAP adaptive
    platform trial
    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
    Illustrates overriding a numeric ordinal outcome (organ
    support-free days) by a binary ordinal outcome (death; this is
    ``sequential composition''), which
  would also be an intercurrent event.}
  \label{fig:remap_cap_override}
\end{figure}

\subsubsection{Increasing precision by reducing the probability of the
largest category}

@@

We have simulation results and some math to show that the value of $\sum_c
\overline p_c^3$ is \emph{essentially a function of the largest category
probability} $\overline{p_{\max}} = \max \overline{p_c}$. See Figure
\ref{fig:p_max} for an illustration.

@@

\subsubsection{Augmenting a state outcome with longitudinal
information}

@@

\subsection{Reasons to coarsen ordinal outcomes}

\subsubsection{Coarsening to avoid directly comparing levels}

@@

\subsubsection{Coarsening for ease of communication}

@@

\section{Choosing ordinal effect measures}

Key references:
\citet{agrestiMeasuresNominalOrdinalAssociation1981,agrestiOrdinalProbabilityEffect2017a,
agrestiSimpleWaysInterpret2018a}

Goal: we want to compare two or more distributions over ordinal
outcomes with a single number. (Two distributions when comparing control and
experimental groups; more distributions when e.g. adjusting for covariates.)

In general this represents a loss of information. Let's say the
outcome has $C$ categories. Then the distribution in each
arm is specified by $C-1$ numbers (for the first $C-1$ probabilities,
  or cumulative logits, or stopping ratio logits--reflecting the
sum-to-1 constraint on probability distributions) and so the
\emph{difference} between the distributions is also specified by
$C-1$ numbers (for example the differences between the cumulative
logits, or logit stopping probabilities).

An effect measure thus represents a way to collapse these $C-1$
numbers into one.

It usually doesn't make sense to consider effect measures based on
numeric coding (e.g. difference in mean levels). This is adding
information that is not present in the basic ordinal structure.

\subsection{Model-based effect measures}

\begin{figure}
  \includegraphics[width=7in]{illustrating_common_or.pdf}
  \caption{The common odds ratio, a model-based effect measure, comes
    from the best-fitting parallel curves to the logit transformed
    empirical cumulative distribution. \textbf{Ideally would illustrate
      using toy bar charts as above, with only 2 groups? Although it's a
  strength that model-based measures can compare multiple groups.}}
  \label{fig:illustrating_or}
\end{figure}

Model-based: common odds ratio; common stopping ratio or common continuation
ratio.

What they have in common: the effect measure is a byproduct of fitting a model
to the data. See Figure \ref{fig:illustrating_or} for an illustration
for the common odds ratio.

The model typically assumes that a certain pattern holds in the outcome
distributions across groups. If the pattern holds, then the differences in
distributions can be described by a single number (with more than two groups,
$G-1$ numbers) instead of $(C-1)$ numbers.

In Figure \ref{fig:illustrating_or} the pattern assumed to hold is
that the logit-transformed eCDFs in each group are parallel.

\subsection{Model-free or nonparametric effect measures}

These are not based on a model (assumed pattern in the differences
between outcome distributions).

\subsubsection{Effect measures based on ``independent pairwise
comparison'' construction}

\begin{figure}
  \includegraphics[width=7.5in]{effect_measures_pp_plot.pdf}
  \caption{Illustrating independent pairwise comparison-based effect
    measures using a PP-plot or ROC curve. Adapted from
  \citet{smithsonReceiverOperatingCharacteristic2023}.}
  \label{fig:illustrating_pairwise_comparisons}
\end{figure}

Probabilistic index/probability of
superiority, win ratio, win odds, etc. These predate win ratio and
DOOR methods
\citep[e.g.][p.~14]{agrestiAnalysisOrdinalCategorical2010}.

Rather they are all be based on the ``\emph{pairs construction}'',
where we imagine independently sampling patients from the control and
treatment groups and considering ``wins'' ``ties'' and ``losses''
(corresponding to concordant and discordant pairs and ties in $y$
respectively). See Figure \ref{fig:illustrating_pairwise_comparisons}
for an illustration.

They can
also be \emph{estimated from data} using pairs of patients, but this
is not essential and they could also be estimated by fitting a
proportional odds model, for example (though they would not be a
simple function of a coefficient).

\subsubsection{Dichotomizing or comparing with a reference group}

\begin{itemize}
  \item Dichotomizing by picking a specific cutpoint,
    e.g. ``hospitalization or
    worse''. This is effectively using a binary outcome, but with the
    ordinal structure used to gain power if fitting a model.
  \item Dichotomizing using quantiles of a reference
    group; for example,
    ``worse than the median outcome under standard care''
  \item Simultaneously comparing all quantiles to those of a
    reference group: ridit analysis
    \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
      smithsonReceiverOperatingCharacteristic2023,
    jansenRiditAnalysisReview1984}
\end{itemize}

\subsection{Comparing model-based and nonparametric effect measures}

Both of these can be considered ``estimands'', insofar as we consider the
estimand associated with the model-based effect measures to be given by the
best-fitting model in the population.

Two classes of assumption violations for model-based effect measures
are \emph{magnitude}-heterogeneity across cutpoints (for example,
  effect of a treatment on hospital length of stay or worse is larger
  than its effect on
death on the relevant scale) and
\emph{sign}-heterogeneity
across cutpoints (failure of stochastic dominance -- for example,
treatment is beneficial for hospital length of stay, harmful for death).

Nonparametric effect measures may appear to be assumption-free. However

\begin{itemize}
  \item Assumptions don't need to hold exactly for the model-based
    effect measures to make sense
  \item In particular, if stochastic dominance holds, we can think of
    both model-based and nonparametric effect measures as
    testing the null hypothesis of no effect ``in a particular
    direction''. A weak interpretation of any particular
    effect measure
    is that it supplies this ``direction''.

    For example the Wilcoxon test is
    equivalent to the score test from fitting a proportional odds model.
    Thus fitting a proportional odds model (or any other model) has a
    valid null-testing
    interpretation even if proportional odds doesn't hold.
  \item Under failure of stochastic dominance (``sign
    heterogeneity''), \emph{neither} parametric nor nonparametric
    effect measures make sense, since it doesn't make sense to
    summarize effects by a single number. Need to dig into the
    cutpoint-specific effects.
  \item Model-free methods don't define a common scale for comparing
    more than two interventions. For example, model-free effect
    measures can be \emph{nontransitive}, so it's not clear how to
    use them in $\geq 2$-arm studies
  \item For the same reason, model-based methods have additional
    advantages, for example
    the possibility of adjustment for covariates.

    When additivity approximately holds, this yields a ``best of both
    worlds'' effect measure that is
    personalized (conditional) yet reportable as a single
    number.
\end{itemize}

Harrell critiques \citep{harrellOverviewCompositeOutcome2024,
harrellRareDegenerativeDiseases2024, harrellViewsCompositeOutcome}:
mainly that
win probability etc. are \emph{relative measures}:

\begin{itemize}
  \item Comparable to Cohen's $d$ or $t$-statistic (difference in
    means divided by standard deviation or standard error) rather than
    difference in means: compare: standardized effect size of 0.2 vs.
    difference in means of 10 mmHg
  \item Affected by study inclusion criteria (why? probably needs to
    be fleshed out.)
\end{itemize}

Simulations show a typically very strong association between
proportional odds ratios and $c$-index
\citep{harrellViolationProportionalOdds2020}.

One point of view is that one would ``ideally'' analyze ordinal
outcomes by reporting the associated expected utilities \todo{cite Berry
talk; example of utility-weighted Rankin scale} and that since
existing ordinal methods supply such utilities implicitly why not
just supply them explicitly. A counterpoint is that the methods do
not, in fact,
supply utilities implicitly.

We should distinguish between effect measures used for \emph{analysis} and for
\emph{reporting}. It may make sense to report the probabilistic index or a
single-cutpoint binary outcome even if the analysis is undertaken using a
proportional odds model.

\section{Situating win/DOOR/etc. methods in the broader context of ordinal
outcomes and effect measures}

This paper is focused on first two steps.

Why is it useful to consider these three steps separately?

\todo{``Partial credit'' analysis?}

@@

\section{Bibliography}

>>>>

\addbibresource{references.bib}
\graphicspath{{fig/}}

\title{Constructing ordinal outcomes for clinical trials}
\author{Leon Di Stefano + Ordinal
Working Group}
\date{\today}

@@

\subsection{Desiderata for (ordinal) outcomes: validity and efficiency}

@@

An example of an outcome which is efficient but which may be invalid
might be continuously-measured concentrations of a biomarker which is a poor
surrogate for disease progression.

One point of view is that one would ``ideally'' analyze ordinal outcomes by
reporting the associated expected utilities \todo{cite Berry talk; example of
utility-weighted Rankin scale} and that since existing ordinal methods supply
such utilities implicitly why not just supply them explicitly. A counterpoint
is that the methods do not, in fact, supply utilities implicitly.
Another is that ordinal outcomes let one ``hedge'' about precise
utillities and this is OK.

\subsection{Constructing versus validating}

\section{Ways to combine ordinal outcomes}

\subsection{Sequential,
hierarchical, \& parallel composition}

@@

@@

\begin{figure}
  \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{Ordinal outcome used in the planned BANDICOOT adaptive
    platform trial \citep{walkerCodesigningNovelOrdinal2025}.
  Illustrates \textbf{parallel composition} of component outcomes.}
  \label{fig:parallel_bandicoot}
\end{figure}

\subsection{Converting back to numeric scores}

This doesn't respect the ordinal nature of the variables. That is, it
implicitly uses information about the \emph{differences} between levels.

\subsection{Pairwise comparison methods (win, DOOR) implicitly construct ordinal
outcomes}

This paper is focused on the first of these steps only.

Why is it useful to consider the three steps separately?

\section{Reasons to combine ordinal outcomes}

\subsection{Providing a more holistic measure of benefit and harm}

\subsection{Incorporating intercurrent events, especially death,
for validity}

@@

\begin{figure}
  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
  \caption{Ordinal outcome used in the REMAP-CAP adaptive
    platform trial
    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
    Illustrates overriding a numeric ordinal outcome (organ
    support-free days) by a binary ordinal outcome (death; this is
    ``sequential composition''), which
    would also be an intercurrent event. This is another example of
  sequential composition.}
  \label{fig:remap_cap_override}
\end{figure}

\subsection{Increasing precision by reducing the probability of the
largest category}

@@

We have simulation results and some maths to show that the value of $\sum_c
\overline p_c^3$ is \emph{essentially a function of the largest category
probability} $\overline{p_{\max}} = \max \overline{p_c}$. See Figure
\ref{fig:p_max} for an illustration of the maths.

@@

\subsection{Augmenting a state outcome with longitudinal
information}

@@

\section{Reasons to coarsen ordinal outcomes}

\subsection{Coarsening to avoid directly comparing levels}

@@

\subsection{Coarsening for ease of communication}

@@

@@
